### 2022-06-15 chapters/interpretability/imli/baseline.tex
imli paper updated with notations

	\begin{equation}
		\label{interpretability_imli_soft_clause_error}
		E_l:=\neg \eta_l;  \qquad \mathsf{wt}(E_l)= 1
	\end{equation}
	Intuitively, if a sample is misclassified, the associated error variable becomes true, thereby dissatisfying the soft clause $ E_l $.

	\item \textbf{Soft clauses for minimizing rule-sparsity:} To favor rule-sparsity, {\imli} tries to learn a classifier with as few literals as possible. Hence, for each feature variable $ \bool_j^i $, {\imli} constructs a unit clause as $ \neg \bool_j^i $. Similar to training accuracy, the weight for this clause is derived as $ \lambda $ from Eq.~\eqref{interpretability_imli_eq:obj}.
	\begin{equation}
		\label{interpretability_imli_soft_clause_feature}
		S_j^i:= \neg \bool_j^i; \qquad \mathsf{wt}(S_j^i)= \lambda
	\end{equation}

@@

	\begin{equation}
			\label{interpretability_imli_hard_clause_rule_consistency}
		H_l:= \neg \eta_l \rightarrow \Big(y^{(l)}\leftrightarrow \bigwedge_{i=1}^k {\mathbf{x}^{(l)}} \circ {\boolset_{i}}\Big); \qquad  \mathsf{wt}(H_l)=\infty
	\end{equation}

>>>>

	\begin{equation}
		\label{interpretability_imli_soft_clause_error}
		E_l:=\neg \eta_l;  \qquad \clauseweight{E_l}= 1
	\end{equation}
	Intuitively, if a sample is misclassified, the associated error variable becomes true, thereby dissatisfying the soft clause $ E_l $.

	\item \textbf{Soft clauses for minimizing rule-sparsity:} To favor rule-sparsity, {\imli} tries to learn a classifier with as few literals as possible. Hence, for each feature variable $ \bool_j^i $, {\imli} constructs a unit clause as $ \neg \bool_j^i $. Similar to training accuracy, the weight for this clause is derived as $ \lambda $ from Eq.~\eqref{interpretability_imli_eq:obj}.
	\begin{equation}
		\label{interpretability_imli_soft_clause_feature}
		S_j^i:= \neg \bool_j^i; \qquad \clauseweight{S_j^i}= \lambda
	\end{equation}

@@

	\begin{equation}
			\label{interpretability_imli_hard_clause_rule_consistency}
		H_l:= \neg \eta_l \rightarrow \Big(y^{(l)}\leftrightarrow \bigwedge_{i=1}^k {\mathbf{x}^{(l)}} \circ {\boolset_{i}}\Big); \qquad  \clauseweight{H_l}=\infty
	\end{equation}

### 2022-06-15 chapters/interpretability/imli/experiment.tex
imli paper updated with notations

	\subfloat{\includegraphics[scale=0.35]{figures/interpretability/imli/dataset_wdbc_cactus_train_val_fit_time}}
	\subfloat{\includegraphics[scale=0.35]{figures/interpretability/imli/dataset_adult_cactus_train_val_fit_time}}\\
	\subfloat{\includegraphics[scale=0.35]{figures/interpretability/imli/dataset_weatherAUS_cactus_train_val_fit_time}}
	\subfloat{\includegraphics[scale=0.35]{figures/interpretability/imli/dataset_labor_cactus_train_val_fit_time}}

>>>>

	\subfloat{\includegraphics[scale=0.4]{figures/interpretability/imli/dataset_wdbc_cactus_train_val_fit_time}}
	\subfloat{\includegraphics[scale=0.4]{figures/interpretability/imli/dataset_adult_cactus_train_val_fit_time}}\\
	\subfloat{\includegraphics[scale=0.4]{figures/interpretability/imli/dataset_weatherAUS_cactus_train_val_fit_time}}
	\subfloat{\includegraphics[scale=0.4]{figures/interpretability/imli/dataset_labor_cactus_train_val_fit_time}}

### 2022-06-15 chapters/interpretability/imli/incremental_learning.tex
tables and figures updated with short caption
preliminaries notations updated
imli paper updated with notations

\section{Incremental Learning with MaxSAT-based formulation}
\label{interpretability_imli_sec:incremental_learning}
The complexity of the MaxSAT query in Section~\ref{interpretability_imli_sec:encoding} increases with the number of samples in the training dataset and the number of clauses to learn in a CNF formula. In this section, we discuss an incremental learning technique that along with MaxSAT formulation can generate classification rules efficiently. Our incremental learning is built on two concepts: (i)
mini-batch learning and (ii) iterative learning. We discuss both in detail in the following.

@@

 where $ C_i $ and $ C'_i $ are the $ i^\text{th} $ clause in $ \Rule $ and $ \Rule' $, respectively and $ \mathds{1} $ is an indicator function that returns $ 1 $ if the input is true and returns $ 0 $ otherwise.  Intuitively, $ d_{H}(\Rule', \Rule) $ calculates the total number of different literals in each (ordered) pair of clauses in $ \Rule $  and $ \Rule' $.  For example, consider $ \Rule = (x_1 \vee x_2) \wedge (\neg x_1) $ and $ \Rule' = (\neg x_1 \vee x_2) \wedge (\neg x_1) $. Then $ 	d_{H}(\Rule, \Rule') = 2 + 0 = 2 $ because in the first clause, the  literals in $ \{x_1, \neg x_1\} $ are absent in either formulas, and the second clause is identical for both $ \Rule $ and $ \Rule' $. In the following, we discuss a modified objective function for mini-batch learning.

\paragraph{Objective Function.}
\label{interpretability_imli_sec:obj_incremental}
In mini-batch learning, we design an objective function to penalize both classification errors on the current batch and the Hamming distance between new and previous classifiers. Let $ (\mathbf{X}^b, \mathbf{y}^b) $ be the current mini-batch. The objective function in mini-batch learning is

%\red{Moreover, $ (\mathbf{X}, \mathbf{y}) $ is the training dataset, on which we measure the performance of $ \Rule $ after completing incremental learning\textemdash we will expand shortly on how we measure the performance. 
%}

\begin{align}
\label{interpretability_imli_eq:obj_incr}
\min_{\Rule} |\mathcal{E}_{\mathbf{X}_b}|+\lambda d_{H}(\Rule, \Rule').
\end{align}

In the objective function, $ \mathcal{E}_{\mathbf{X}_b} $ is the misclassified subset of samples in the current batch $ (\mathbf{X}^b, \mathbf{y}^b) $. Unlike controlling the rule-sparsity in the non-incremental approach in the earlier section, in Eq.~\eqref{interpretability_imli_eq:obj_incr}, $ \lambda $ controls the trade-off between classification errors and the syntactic differences between consecutive classifiers. We next discuss how to encode the modified objective function as a MaxSAT query.

\paragraph{MaxSAT Encoding of Mini-batch Learning.}
\label{interpretability_imli_sec:encoding_incremental}
In order to account for the modified objective function, we only modify the soft clause $ S_j^i $ in the MaxSAT query $ Q $. In particular, we define $ S_j^i $ to penalize for the complemented assignment  of the feature variable $ b_j^i $ in $ \Rule $ compared to $ \Rule' $.
\[
S_j^i:=
\begin{cases}
b_j^i& \text{if }  x_j \in  \clause(\Rule',i)  \\
\neg b_j^i & \text{otherwise}
\end{cases};\qquad \mathsf{wt}(S_j^i)=\lambda
\]

$ S_j^i $ is either a unit clause $ b_j^i $ or $ \neg b_j^i $ depending on whether the associated feature $ x_j $ appears in the previous classifier $ \Rule' $ or not. In this encoding, the Hamming distance of $ \Rule $ and $ \Rule' $ is minimized while attaining minimum classification errors on the current batch (using soft clause $ E_l $ in Eq.~\eqref{interpretability_imli_soft_clause_error}) To this end,  mini-batch learning starts with an empty CNF formula as $ \Rule' $ with no feature, and thus $ S_j^i := \neg b_j^i $ for the first batch. We next analyze the complexity of the MaxSAT encoding for mini-batch learning.

\begin{proposition}
	\label{interpretability_imli_prop:maxsat_complexity_incremental}
	Let $ n' \triangleq |\mathbf{X}_b| \ll |\mathbf{X}| $ be the size of a mini-batch, $ n'_{neg} \le n' $ be the number of negative samples in the batch, and $ m $ be the number of Boolean features. According to Proposition~\ref{interpretability_imli_prop:maxsat_variables}, to learn a $ k $-clause CNF classifier in mini-batch learning, the MaxSAT encoding for each batch has $ km + n' $ Boolean variables and $ kn'_{neg} $ auxiliary variables. Let $ n'_{pos} = n' - n'_{neg} $ be the number of positive samples in the batch. Then, according to Proposition~\ref{interpretability_imli_prop:maxsat_clauses}, the number of clauses in the MaxSAT query is $ k m+n'+k n'_{pos}+(k m+1)n'_{neg} \approx \mathcal{O}(k m  n' ) $ when $ n'_{pos}= n'_{neg} =\frac{n'}{2} $.
\end{proposition}

@@

\paragraph{Assessing  the Performance of $ \Rule $.} Since we apply a heuristic objective in mini-batch learning,  $ \Rule $ may be optimized for the current batch while generalizing poorly on the full training data. To tackle this, after learning on each batch, we decide whether to keep $ \Rule $ or not by assessing the performance of $ \Rule $ on the training data and keep $ \Rule $ if it achieves higher performance. We measure the performance of $ \Rule $ on the full training data $ (\mathbf{X}, \mathbf{y}) $ using a weighted combination of classification errors and rule-size. In particular, we compute a combined loss function $ \mathsf{loss}(\Rule) \triangleq |\mathcal{E}_\mathbf{X}| +\lambda |\mathcal{R}| $ on the training data $ (\mathbf{X}, \mathbf{y}) $, which is indeed the value of the objective function that we minimize in the non-incremental learning in Section~\ref{interpretability_imli_sec:problem}. Additionally, we  discard the current classifier $ \Rule $ when the loss does not decrease ($ \mathsf{loss}(\Rule) > \mathsf{loss}(\Rule') $).

We present the algorithm for mini-batch learning in Algorithm~\ref{interpretability_imli_algo:incremental}. 
\begin{algorithm}
	\caption{MaxSAT-based Mini-batch Learning}
	\label{interpretability_imli_algo:incremental}
	\begin{algorithmic}[1]
		\Function{$ \mathsf{MiniBatchLearning} $}{$ \mathsf{X},\mathbf{y},\lambda, k $}
		\State $\mathcal{R} = \leftarrow \bigwedge_{i=1}^k\text{false} $ \Comment{Empty CNF formula}
		\State $ \mathsf{loss}_{\max} \leftarrow \infty$ 
		\For{$ i \leftarrow 1, \dots, N $} \Comment{$ N $ is the total batch-count}
		\State $ \mathbf{X}^b, \mathbf{y}^b \leftarrow \mathsf{GetBatch}(\mathsf{X},\mathbf{y}) $
		\State $ Q, \weight{\cdot} \leftarrow \mathsf{MaxSATEncoding}(\Rule, \mathbf{X}^b, \mathbf{y}^b, \lambda, k) $ \Comment{Returns a weighted-partial CNF}
		\State $ \sigma^* \leftarrow \MaxSAT(Q,\weight{\cdot}) $
		\State $ \Rule_{\mathsf{new}} \leftarrow \mathsf{ConstructClassifier(\sigma^*)} $
		\State $ \mathsf{loss} \leftarrow  |\mathcal{E}_\mathbf{X}| +\lambda |\mathcal{R}_{\mathsf{new}}| $ \Comment{Compute loss}
		\If{$ \mathsf{loss} < \mathsf{loss}_{\max} $}
		\State $ \mathsf{loss}_{\max} \leftarrow \mathsf{loss} $
		\State $ \Rule \leftarrow \Rule_{\mathsf{new}} $
		\EndIf
		\EndFor
		\State \Return $ \Rule $
		\EndFunction
	\end{algorithmic}
\end{algorithm}

>>>>

\section{Incremental Learning with MaxSAT-based formulation}
\label{interpretability_imli_sec:incremental_learning}
The complexity of the MaxSAT query in Section~\ref{interpretability_imli_sec:encoding} increases with the number of samples in the training dataset and the number of clauses to learn in a CNF formula. In this section, we discuss an incremental learning technique that along with the MaxSAT formulation can generate classification rules efficiently. Our incremental learning is built on two concepts: (i)
mini-batch learning and (ii) iterative learning. We discuss both in detail in the following.

@@

 where $ C_i $ and $ C'_i $ are the $ i^\text{th} $ clause in $ \Rule $ and $ \Rule' $, respectively and $ \mathds{1} $ is an indicator function that returns $ 1 $ if the input is true and returns $ 0 $ otherwise.  Intuitively, $ d_{H}(\Rule', \Rule) $ calculates the total number of different literals in each (ordered) pair of clauses in $ \Rule $  and $ \Rule' $.  For example, consider $ \Rule = (X_1 \vee X_2) \wedge (\neg X_1) $ and $ \Rule' = (\neg X_1 \vee X_2) \wedge (\neg X_1) $. Then $ 	d_{H}(\Rule, \Rule') = 2 + 0 = 2 $ because in the first clause, the  literals in $ \{X_1, \neg X_1\} $ are absent in either formulas, and the second clause is identical for both $ \Rule $ and $ \Rule' $. In the following, we discuss a modified objective function for mini-batch learning.

\paragraph{Objective Function.}
\label{interpretability_imli_sec:obj_incremental}
In mini-batch learning, we design an objective function to penalize both classification errors on the current batch and the Hamming distance between new and previous classifiers. Let $ \dataset_b \subset \dataset $ be the current mini-batch, where $ |\dataset_b| \ll |\dataset| $. The objective function in mini-batch learning is

%\red{Moreover, $ (\red{\dataset}, \mathbf{y}) $ is the training dataset, on which we measure the performance of $ \Rule $ after completing incremental learning\textemdash we will expand shortly on how we measure the performance. 
%}

\begin{align}
\label{interpretability_imli_eq:obj_incr}
\min_{\Rule} |\mathcal{E}_{\dataset_b}|+\lambda d_{H}(\Rule, \Rule').
\end{align}

In the objective function, $ \mathcal{E}_{\dataset_b} $ is the misclassified subset of samples in the current batch $ \dataset_b $. Unlike controlling the rule-sparsity in the non-incremental approach in the earlier section, in Eq.~\eqref{interpretability_imli_eq:obj_incr}, $ \lambda $ \emph{controls the trade-off between classification errors and the syntactic differences between consecutive classifiers}. We next discuss how to encode the modified objective function as a MaxSAT query.

\paragraph{MaxSAT Encoding of Mini-batch Learning.}
\label{interpretability_imli_sec:encoding_incremental}
In order to account for the modified objective function, we only modify the soft clause $ S_j^i $ in the MaxSAT query $ Q $. In particular, we define $ S_j^i $ to penalize for the complemented assignment  of the feature variable $ \bool_j^i $ in $ \Rule $ compared to $ \Rule' $.
\[
S_j^i:=
\begin{cases}
\bool_j^i& \text{if }  X_j \in  \clause(\Rule',i)  \\
\neg \bool_j^i & \text{otherwise}
\end{cases};\qquad \clauseweight{S_j^i}=\lambda
\]

$ S_j^i $ is either a unit clause $ \bool_j^i $ or $ \neg \bool_j^i $ depending on whether the associated feature $ X_j $ appears in the previous classifier $ \Rule' $ or not. In this encoding, the Hamming distance of $ \Rule $ and $ \Rule' $ is minimized while attaining minimum classification errors on the current batch (using soft clause $ E_l $ in Eq.~\eqref{interpretability_imli_soft_clause_error}) To this end,  mini-batch learning starts with an empty CNF formula as $ \Rule' $ with no feature, and thus $ S_j^i := \neg \bool_j^i $ for the first batch. We next analyze the complexity of the MaxSAT encoding for mini-batch learning.

\begin{proposition}
	\label{interpretability_imli_prop:maxsat_complexity_incremental}
	Let $ n' \triangleq |\dataset_b| \ll |\dataset| $ be the size of a mini-batch, $ n'_{neg} \le n' $ be the number of negative samples in the batch, and $ m $ be the number of Boolean features. According to Proposition~\ref{interpretability_imli_prop:maxsat_variables}, to learn a $ k $-clause CNF classifier in mini-batch learning, the MaxSAT encoding for each batch has $ km + n' $ Boolean variables and $ kn'_{neg} $ auxiliary variables. Let $ n'_{pos} = n' - n'_{neg} $ be the number of positive samples in the batch. Then, according to Proposition~\ref{interpretability_imli_prop:maxsat_clauses}, the number of clauses in the MaxSAT query is $ k m+n'+k n'_{pos}+(k m+1)n'_{neg} \approx \mathcal{O}(k m  n' ) $ when $ n'_{pos}= n'_{neg} =\frac{n'}{2} $.
\end{proposition}

@@

\paragraph{Assessing  the Performance of $ \Rule $.} Since we apply a heuristic objective in mini-batch learning,  $ \Rule $ may be optimized for the current batch while generalizing poorly on the full training data. To tackle this, after learning on each batch, we decide whether to keep $ \Rule $ or not by assessing the performance of $ \Rule $ on the training data and keep $ \Rule $ if it achieves higher performance. We measure the performance of $ \Rule $ on the full training data $ \dataset $ using a weighted combination of classification errors and rule-size. In particular, we compute a combined loss function $ \mathsf{loss}(\Rule) \triangleq |\mathcal{E}_\dataset| +\lambda |\mathcal{R}| $ on the training data $ \dataset $, which is indeed the value of the objective function that we minimize in the non-incremental learning in Section~\ref{interpretability_imli_sec:problem}. Additionally, we  discard the current classifier $ \Rule $ when the loss does not decrease ($ \mathsf{loss}(\Rule) > \mathsf{loss}(\Rule') $).

We present the algorithm for mini-batch learning in Algorithm~\ref{interpretability_imli_algo:incremental}. 
\begin{algorithm}
	\caption{MaxSAT-based Mini-batch Learning}
	\label{interpretability_imli_algo:incremental}
	\begin{algorithmic}[1]
		\Function{MiniBatchLearning}{$ \dataset, \lambda, k $}
		\State $\mathcal{R} = \leftarrow \bigwedge_{i=1}^k\text{false} $ \Comment{Empty CNF formula}
		\State $ \mathsf{loss}_{\max} \leftarrow \infty$ 
		\For{$ i \leftarrow 1, \dots, N $} \Comment{$ N $ is the total batch-count}
		\State $ \dataset_b \leftarrow \textsc{GetBatch}(\dataset) $
		\State $ Q, \clauseweight{\cdot} \leftarrow \textsc{MaxSATEncoding}(\Rule, \dataset_b, \lambda, k) $ \Comment{Returns a weighted-partial CNF}
		\State $ \sigma^* \leftarrow \MaxSAT(Q,\clauseweight{\cdot}) $
		\State $ \Rule_{\mathsf{new}} \leftarrow \textsc{ConstructClassifier}(\sigma^*) $
		\State $ \mathsf{loss} \leftarrow  |\mathcal{E}_\dataset| +\lambda |\mathcal{R}_{\mathsf{new}}| $ \Comment{Compute loss}
		\If{$ \mathsf{loss} < \mathsf{loss}_{\max} $}
		\State $ \mathsf{loss}_{\max} \leftarrow \mathsf{loss} $
		\State $ \Rule \leftarrow \Rule_{\mathsf{new}} $
		\EndIf
		\EndFor
		\State \Return $ \Rule $
		\EndFunction
	\end{algorithmic}
\end{algorithm}
